---
html_document:
  code_folding: hide
  number_sections: yes
  self_contained: yes
  toc: yes
  toc_float: yes
author: "Dobgima Mofor"
date: "2024-01-11"
pdf_document:
  toc: yes
title: "Epi measures"
editor_options:
  markdown:
    wrap: 72
---

```{r results='hide', message=FALSE, warning=FALSE, echo=FALSE}

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```


```{r, warning=FALSE}
library(readr) # Import Excel files
library(tidyverse) #Data manipulation
library(epitools)
```


## I. Basic statistical test in R

### 1. One sample T tests in R

One sample Z and T tests are used for determining whether the mean in a population is different than a hypothesized value

Assumptions for Z and T tests

Large sample size or data are approximately normal if sample size is small

Assumptions for Z test

Population standard deviation is known

_Example_
Is average gestational age in the population different than 39 weeks (use alpha = 0.05)

- Null hypothesis H0 : mu = 39
- Alternative hypothesis H1: mu != 39

```{r}
vlbw <- read_csv('vlbw.csv')

vlbw <- vlbw[complete.cases(vlbw), c('bwt', 'gest', 'sex', 'twn', 'pneumo')]

attach(vlbw)
```


```{r}
a <- hist(vlbw$gest, xlab = 'Gestational age', main = 'Histogram gestational age')

b <- ggplot(vlbw, aes(gest)) +
  geom_histogram(col = 1, fill = 2, bins = 10) +
  theme_classic()
a
b
#library(ggpubr)
#ggarrange(a,b)
```

Perform a t-test for
H0: mu = 39
H1: mu != 39
at alpha = 0.05

```{r}
t_age <- t.test(vlbw$gest, mu = 39)
names(t_age)
```

### 2. Two sample t-tests

Unpaired two sample t-tests

Recall that a two sample t-test tests the hypothesis that the means in two populations are the same

i.e Whether the means of a continuous variable differ between two groups:

Null hypothesis H0: mu1 = mu2
Alternative hypothesis H1: mu1 != mu2


_Example_
Does mean gestational age differ between male and female infants?

```{r}
age_female <- vlbw %>% filter(sex == 'female') %>% select(gest)

age_male <- vlbw %>% filter(sex == 'male') %>% select(gest)

t_age_sex <- t.test(age_male, age_female)
t_age_sex$p.value
```
p < 0.05 i.e there is no significant difference between male and female age

- Difference in means > 0

```{r}
t.test(age_male, age_female, alternative = 'less')
```

### 3. Chi-squared test

Are two categorical variables independent?

Hypothesis:

Null hypothesis: Sex is independent of being a twin
Alternative hypothesis: Sex is not independent of being a twin

Assumptions:

- If 2x2 table, no cell counts < 5
- If rxc table, no more than 20% cells < 5

```{r}
chisq.test(sex, twn)
```

# II. Epi measures
# 1. Relative risk (RR) and Odds ratio (OR)


- Ratio of risks: p1/p2
-Is the risk of disease the same in the exposed and unexposed groups?
- Often interested in testing H0 : RR = 1 vs. H1 : RR != 1
- Can only be calculated in prospective studies

```{r}
table_pneumo <- table(pneumo,twn)

epi_pneumo <- epitab(table_pneumo, method = 'riskratio')
epi_pneumo$tab
```

- Ratio of odds
- Is the odds of disease the same in the exposed and unexposed groups?
- Odds is NOT the same as risk
- Odds: p/(1-p) or p/q
- OR = (p/1-p)) / (p2/(1-p2)) = (p1/q1)/ (p2/q2)
- Often interested in testing H0: OR = 1 vs H1: OR != 1
- Useful in retrospective studies

```{r}
epi_odds <- epitab(table_pneumo, method = 'oddsratio')
epi_odds$tab[,c('oddsratio', 'lower', 'upper')]
```
## 2. Sample size calculations
## 3. Power

Power is the ability to detect a difference if one exists

How many observations would we need to test whether two means are different if

- The difference in means is 0.1
- The standard deviation is 1
- We want 90% power

For t tests

```{r}
power.t.test(delta = 0.1, power = 0.9, type = 'two.sample', alternative = 'two.sided')
```

